Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2023-12-09 report/appendix_prepost_constr.Rmd
Reorganised manuscript

The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}. Transition durations were shorter at before-word locations for the post-test. Evidence was negligible for any other comparisons.

@@

file <- list.files(path = "../stanout/plantra", 
                   pattern = "mogbetaun.*.csv$",
                   full.names = T)

@@

write_csv(tmp, "tables/posttest_effect_cellmeans.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for post-test effect. Cell means are shown for the pre-test and post-test in msecs for fluent key-transitions, the slowdown for long transitions and the probability of hesitant transitions. The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Fluent transitions" = 1:3, 
            "Hesitation duration" = 4:6,
            "Hesitation probability" = 7:9))

>>>>

The pre-post test effect for the PLanTra dataset is reported in Table \ref{tab:retesteffect}. Evidence for all comparisons was negligible.

@@

file <- list.files(path = "../stanout/plantra", 
                   pattern = "mogbetaco.*.csv$",
                   full.names = T)

@@

write_csv(tmp, "tables/posttest_effect_cellmeans_constr.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for post-test effect. Cell means are shown for the pre-test and post-test in msecs for the slowdown for hesitant transitions and the probability of hesitant transitions. The effect for post-test is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Hesitation duration" = 1:3,
            "Hesitation probability" = 4:6))

### 2023-12-09 report/appendix_shift.Rmd
Reorganised manuscript

## Key-combination effect (SPL2)

@@

Therefore we compared whether the different patterns can be explain on the basis of the additional keystroked involved in before-sentence transitions. We compared the SPL2 data including and excluding the keystroke after shift. Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations. The results of this comparison can be found in Table \ref{tab:shiftcellmeans}. Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift. Fluent key transitions were substantially longer when including the interval following the shift key. The slowdown for hesitations was affected too but the difference is numerically small. There was no conclusive evidence for an increased hesitation probability.

```{r shiftcellmeans, results = 'asis'}
# cell means and data set difference
cellmeans <- ps_spl2 %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  select(-ds, -group) %>%
  mutate(across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(lang, location, data), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est)

@@

>>>>

## Key-combination effect (unconstrained mixture model)

@@

Therefore we compared whether the different patterns can be explain on the basis of the additional keystroke involved in before-sentence transitions. We compared the SPL2 data including and excluding the keystroke after shift. Although we modelled all transition locations, we present only before-sentence transitions below as there was, as one would expect, no difference at word locations. The results of this comparison can be found in Table \ref{tab:shiftcellmeans2}.

```{r shiftcellmeans2, results = 'asis'}
# cell means and data set difference
cellmeans <- ps_spl2 %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  select(-ds, -group) %>%
  mutate(across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(lang, location, data), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est)

@@

Overall, fluent transition duration and the hesitation duration were affected by whether or not the sentence-initial transition include the character following shift. Fluent key transitions were substantially longer when including the interval following the shift key. The slowdown for hesitations was affected too but the difference is numerically small. There was no conclusive evidence for an increased hesitation probability.

### 2023-12-09 report/appendix_shift_constr.tex
Reorganised manuscript

## Key-combination effect (constrained mixture model)

The analysed datasets differ to the extent that keystroke intervals at before sentence location sometimes did (PLanTra, LIFT) or did not (CATO, C2L1, SPL2, GUNNEXP2) scope over the character following the shift key. In other words, the pause before sentences summed across two key intervals in the PLanTra and LIFT data, namely `_^[shift]^C` but only involved one key interval, namely `_^[shift]` for the remaining datasets. Therefore, longer or more frequent pauses at before-sentence locations compared to before-word locations can be explained without reference to linguistic edges.

<!-- Also there is a possibility that some inconsistencies in our findings can be explained on the basis of including the keystroke following shift.  -->

@@

```{r shiftcellmeans, results = 'asis'}
# cell means and data set difference
cellmeans <- ps_spl2_both %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  select(-ds, -group) %>%
  mutate(across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(lang, location, data), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est) %>% 
  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
  unique()

file1 <- list.files(path = "../stanout/spl2", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

file2 <- list.files(path = "../stanout/spl2_shift", 
                    pattern = "mogbetacon.+.csv$",
                    full.names = T)

tmp <- map_dfr(c(file1, file2), ~read_csv(.x) %>% 
                 mutate(data = .x)) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value))) %>% 
  mutate(across(lang, recode, "EN" = "L1", "ES" = "L2"),
         across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .))) %>% 
  pivot_wider(names_from = data, values_from = value) %>% 
  unnest(c(spl2, spl2_shift)) %>% 
  mutate(diff = spl2_shift - spl2) %>% 
  summarise(across(diff, list(mean = mean, 
                              lower = lower, 
                              upper = upper,
                              BF = BF), 
                   .names = "{.fn}"),
            .by = c(lang, location, param)) %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2)))),
         across(param, recode, theta = "prob")) %>% 
  left_join(cellmeans) %>% 
  arrange(param, location, lang) %>% 
  filter(location %in% c("overall", "before sentence")) %>% 
  select(param, 
         Language = lang,
         `_^[shift] + C` = ends_with(" + C)"),
         `_^[shift]` = ends_with("FT])"),
         "Difference" = mean, BF)

write_csv(tmp, "tables/shift_effect_cellmeans_constr.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for key transitions immediately preceding a sentence. Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions. The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "footnotesize",
          longtable = T,
          stub_indents = list("Fluent transitions" = 1,
                              "Hesitation duration" = 2:3,
                              "Hesitation probability" = 4:5) )

```

The fluent transition duration and hesitation duration were not affected by whether or not the sentence-initial transition include the character following shift for neither language. However, we found strong evidence for an increased hesitation probability -- in both -- languages when the before-sentence key transition included the character following the shift key. Notably, the hesitation probability increased ceiling. It is not entirely surprised that the constrained mixture model identifies essentially all before-sentence transition as pauses: the constrained model does not distinguish between transition locations for the duration of fluent transitions. Therefore essentially all before-sentence transitions are identified as hesitation as they always include two keystrokes while all other transition location (which is the majority of the data) do not.

>>>>

\clearpage
\makeatletter
\efloat@restorefloats
\makeatother

\begin{appendix}
\section{}
\hypertarget{key-combination-effect-constrained-mixture-model}{%
\subsection{Key-combination effect (constrained mixture
model)}\label{key-combination-effect-constrained-mixture-model}}

The analysed datasets differ to the extent that keystroke intervals at
before sentence location sometimes did (PLanTra, LIFT) or did not (CATO,
C2L1, SPL2, GUNNEXP2) scope over the character following the shift key.
In other words, the pause before sentences summed across two key
intervals in the PLanTra and LIFT data, namely
\texttt{\_\^{}{[}shift{]}\^{}C} but only involved one key interval,
namely \texttt{\_\^{}{[}shift{]}} for the remaining datasets. Therefore,
longer or more frequent pauses at before-sentence locations compared to
before-word locations can be explained without reference to linguistic
edges.

@@

\begin{center}
\begin{ThreePartTable}

\begin{TableNotes}[para]
\normalsize{\textit{Note.} PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.}
\end{TableNotes}

\footnotesize{

\begin{longtable}{lrrrr}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:shiftcellmeans}Mixture model estimates for key transitions immediately preceding a sentence. Cell means are shown for transitions that do and do not involve the transition to the character following shift in msecs for the slowdown for long transitions and the probability of hesitant transitions. The difference for including the transition duration to the character after shift is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95\% PIs in brackets.}\\
\toprule
Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:shiftcellmeans} continued}}\\
\toprule
Language & \multicolumn{1}{c}{\_\textasciicircum{}[shift] + C} & \multicolumn{1}{c}{\_\textasciicircum{}[shift]} & \multicolumn{1}{c}{Difference} & \multicolumn{1}{c}{BF}\\
\midrule
\endhead
Fluent transitions &  &  &  & \\
\ \ \ overall & 156 [145, 169] & 159 [147, 172] & -0.02 [-0.13, 0.09] & 0.06\\
Hesitation duration &  &  &  & \\
\ \ \ L1 & 1,167 [1,037, 1,305] & 1,281 [1,082, 1,496] & -0.06 [-0.21, 0.09] & 0.11\\
\ \ \ L2 & 2,019 [1,772, 2,292] & 1,802 [1,557, 2,073] & 0.12 [-0.02, 0.27] & 0.29\\
Hesitation probability &  &  &  & \\
\ \ \ L1 & 1.00 [.99, 1.00] & .78 [.69, .86] & 4.91 [3.68, 6.35] & > 100\\
\ \ \ L2 & 1.00 [.99, 1.00] & .92 [.87, .96] & 3.4 [2.12, 4.85] & > 100\\
\bottomrule
\addlinespace
\insertTableNotes
\end{longtable}

}

\end{ThreePartTable}
\end{center}

The fluent transition duration and hesitation duration were not affected
by whether or not the sentence-initial transition include the character
following shift for neither language. However, we found strong evidence
for an increased hesitation probability -- in both -- languages when the
before-sentence key transition included the character following the
shift key. Notably, the hesitation probability increased ceiling. It is
not entirely surprised that the constrained mixture model identifies
essentially all before-sentence transition as pauses: the constrained
model does not distinguish between transition locations for the duration
of fluent transitions. Therefore essentially all before-sentence
transitions are identified as hesitation as they always include two
keystrokes while all other transition location (which is the majority of
the data) do not.
\end{appendix}

### 2023-12-09 report/appendix_sim.Rmd
Reorganised manuscript

A possible concern with these results -- substantially better predictive performance for bi-modal mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit. We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models. To address this concern we repeated a comparison for a uni-modal model and a bi-modal mixture model for two sets of simulated data that were simulated either with a uni-modal or bi-modal process at heart. In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- uni-modal vs bi-modal -- and we can test whether these models can successfully uncover the true parameter value.

In particular, the first data set was simulated using a bi-modal mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.

@@

The second data set was generated with a uni-modal log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.

@@

Again, this model is a simplified version of the uni-modal models used in the main analysis. The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.

@@

```{r simdata, fig.cap= "Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
files <- list.files("../mixture-model-sim/data", full.names = T)

map_dfr(files, ~read_csv(.) %>% 
        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
  mutate(across(Dataset, ~recode(., mogdata = "bi-modal", lmdata = "uni-modal"))) %>% 
  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
  geom_density(alpha = .25) +
  scale_colour_colorblind() + 
  scale_fill_colorblind() +
  scale_x_log10(labels = ~format(., big.mark = ",")) +
  theme(text = element_text(size = 10)) +
  labs(colour = "Data process", 
       fill = "Data process", 
       x = "y", 
       y = "Density")
```

For each of these two data sets we simulated 1,000 observations. We run 2 models -- a bi-modal mixture model and a uni-modal model -- each for both data sets. Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup. Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}. The parameters are shown by type of data generating process along with the true parameter values. Parameter value estimates are shown by type of Bayesian model. The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process. Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they uni-modal model.

@@

names(table) <- c("Model", "Parameter with true value", "Bi-modal model", "Uni-modal model")

apa_table(table[,-1], 
          escape = FALSE,
      col_spanners = list("Estimate with 95\\% PI" = 2:3),
      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
      align = c("l", rep("r", 2)),
      stub_indents = list(`Bi-modal data` = 1:5, 
                          `Uni-modal data` = 6:7))

@@

We used LOO-CV to compare the predictive performance of the two models for each data generating process. The model comparisons can be found for each data generating process in Table \ref{tab:loossim}. The results rule out the possibility that the mixture model does always lead to higher predictive performance. Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a uni-modal process. However, for the data generated with a bi-modal process, the mixture model shows a substantially higher predictive performance. In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error, as metric for the strength of evidence (Sivula et al., 2020), shows that the mixture model performs 11.6 times better than the uni-modal model. In comparison, for the uni-modal data, the uni-modal model performs only 0.77 times better than the bi-modal mixture model. Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the uni-modal is preferred by the law of parsimony.

```{r loossim, results='asis'}
table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
  select(1:6) %>% 
  mutate(
#    elpd_se_ratio = abs(elpd_diff / se_diff),
    across(ends_with("diff"), ~round(., 1)),
    across(ends_with("loo"), ~round(., 0)),
    across(where(is.numeric), ~format(., big.mark = ",")),
    across(everything(), ~as.character(.)),
    across(everything(), ~str_trim(.)),
    across(everything(), ~str_replace_all(., "NaN", "--")),
    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
    across(c(data, model), ~recode(., mog = "Bi-modal mixture model",
                                      lm = "Uni-modal model"))) %>% 
  select(data, model, starts_with("elpd"))

@@

apa_table(table[,-1], 
          escape = FALSE,
      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
      align =c("l", rep("r", 2)),
      stub_indents = list(`Data: Bi-modal mixture process` = 1:2, 
                          `Data: Uni-modal process` = 3:4))

>>>>

A possible concern with these results -- substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit. We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models. To address this concern we repeated a comparison for a single distribution model and a two-distrobutionsl mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart. In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.

In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equations in \ref{eq:bimodcon} and \ref{eq:bimoduncon}). This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.

@@

The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.

@@

Again, this model is a simplified version of the single distribution models used in the main analysis. The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.

@@

```{r simdata, fig.cap= "Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
files <- list.files("../mixture-model-sim/data", full.names = T)

map_dfr(files, ~read_csv(.) %>% 
        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
  mutate(across(Dataset, ~recode(., mogdata = "two distributions", lmdata = "single distributions"))) %>% 
  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
  geom_density(alpha = .25, 
               linewidth = .25) +
  scale_colour_colorblind() + 
  scale_fill_colorblind() +
  scale_x_log10(labels = ~format(., big.mark = ",")) +
  theme(text = element_text(size = 10)) +
  labs(colour = "Data process", 
       fill = "Data process", 
       x = "y", 
       y = "Density")
```

For each of these two data sets we simulated 1,000 observations. We run 2 models -- a two-distribution mixture model and a single-distribution model -- each for both data sets. Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup. Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}. The parameters are shown by type of data generating process along with the true parameter values. Parameter value estimates are shown by type of Bayesian model. The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process. Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they single-distribution model.

@@

names(table) <- c("Model", "Parameter with true value", "Two-distributions model", "Single-distributions model")

apa_table(table[,-1], 
          escape = FALSE,
      col_spanners = list("Estimate with 95\\% PI" = 2:3),
      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
      align = c("l", rep("r", 2)),
      stub_indents = list(`Two-distributions data` = 1:5, 
                          `Single-distribution data` = 6:7))

@@

We used LOO-CV to compare the predictive performance of the two models for each data generating process. The model comparisons can be found for each data generating process in Table \ref{tab:loossim}. The results rule out the possibility that the mixture model does always lead to higher predictive performance. Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a single-distribution process. However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance. In fact, the ratio of $\Delta\widehat{elpd}$ and its standard error -- a *z*-score approximation -- reveals a value of 11.6 which is strong evidence in in favour of the mixture model compared to the single-distribution model (Sivula et al., 2020). In comparison, for the single-distribution data, the single-distribution model was negligibly better than the two-distribution mixture model which was evidence by a value of 0.77 for the ratio of $\Delta\widehat{elpd}$ and its standard error. Thus, as the difference $\Delta\widehat{elpd}$ between the model is negligible, the single distribution is preferred by the law of parsimony.

```{r loossim, results='asis'}
table <- read_csv("../mixture-model-sim/stanout/modelcomparison.csv") %>% 
  select(1:6) %>% 
  mutate(
#    elpd_se_ratio = abs(elpd_diff / se_diff),
    across(ends_with("diff"), ~round(., 1)),
    across(ends_with("loo"), ~round(., 0)),
    across(where(is.numeric), ~format(., big.mark = ",")),
    across(everything(), ~as.character(.)),
    across(everything(), ~str_trim(.)),
    across(everything(), ~str_replace_all(., "NaN", "--")),
    across(elpd_diff, ~str_c(., " (", se_diff, ")")),
    across(elpd_loo, ~str_c(., " (", se_elpd_loo, ")")),
    across(elpd_diff, ~str_replace_all(., "0.0 \\(0.0\\)", "--")),
    across(c(data, model), ~recode(., mog = "Two-distribution mixture model",
                                      lm = "Single-distribution model"))) %>% 
  select(data, model, starts_with("elpd"))

@@

apa_table(table[,-1], 
          escape = FALSE,
      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
      align =c("l", rep("r", 2)),
      stub_indents = list(`Data: Two-distribution mixture process` = 1:2, 
                          `Data: Single-distribution process` = 3:4))

### 2023-12-09 report/appendix_sim.tex
Reorganised manuscript

Again, this model is a simplified version of the uni-modal models used
in the main analysis. The model assumes a log-Gaussian distribution with
a mean parameter \(\beta\) and a standard deviation \(\sigma^2\).

@@

\caption{Data simulated with a bi-modal (grey) and a uni-modal (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}(\#fig:simdata)
\end{figure}

For each of these two data sets we simulated 1,000 observations. We run
2 models -- a bi-modal mixture model and a uni-modal model -- each for
both data sets. Models were run with 3 chains, with each 6,000
iterations of which 3,000 were warmup. Estimates with 95\% probability
intervals are shown in Table \ref{tab:simparam}. The parameters are
shown by type of data generating process along with the true parameter
values. Parameter value estimates are shown by type of Bayesian model.
The results show that either of the two Bayesian models succesfully
uncovered the model parameters of the data with its corresponding data
generating process, as shown in Table \ref{tab:simparam}, but less so
when the model was applied to data generated with the incorrect
underlying process. Particularly the mixing proportion \(\theta\) and
the slowdown parameter \(\delta\) were not uncovered at all by they
uni-modal model.

@@

\begin{tabular}{lrr}
\toprule
 & \multicolumn{2}{c}{Estimate with 95\% PI} \\
\cmidrule(r){2-3}
Parameter with true value & \multicolumn{1}{c}{Bi-modal model} & \multicolumn{1}{c}{Uni-modal model}\\
\midrule
Bi-modal data &  & \\
\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
Uni-modal data &  & \\
\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
\bottomrule
\end{tabular}

@@

We used LOO-CV to compare the predictive performance of the two models
for each data generating process. The model comparisons can be found for
each data generating process in Table \ref{tab:loossim}. The results
rule out the possibility that the mixture model does always lead to
higher predictive performance. Indeed, the mixture model showed a
slightly lower predictive performance for the data that were generated
with a uni-modal process. However, for the data generated with a
bi-modal process, the mixture model shows a substantially higher
predictive performance. In fact, the ratio of \(\Delta\widehat{elpd}\)
and its standard error, as metric for the strength of evidence (Sivula
et al., 2020), shows that the mixture model performs 11.6 times better
than the uni-modal model. In comparison, for the uni-modal data, the
uni-modal model performs only 0.77 times better than the bi-modal
mixture model. Thus, as the difference \(\Delta\widehat{elpd}\) between
the model is negligible, the uni-modal is preferred by the law of
parsimony.

@@

\begin{tabular}{lrr}
\toprule
Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
\midrule
Data: Bi-modal mixture process &  & \\
\ \ \ Bi-modal mixture model & -- & -6,068 (41)\\
\ \ \ Uni-modal model & -191.3 (16.5) & -6,259 (38)\\
Data: Uni-modal process &  & \\
\ \ \ Uni-modal model & -- & -5,030 (24)\\
\ \ \ Bi-modal mixture model & -0.5 (0.7) & -5,030 (24)\\
\bottomrule
\addlinespace
\end{tabular}

>>>>

Again, this model is a simplified version of the single distribution
models used in the main analysis. The model assumes a log-Gaussian
distribution with a mean parameter \(\beta\) and a standard deviation
\(\sigma^2\).

@@

\caption{Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.}(\#fig:simdata)
\end{figure}

For each of these two data sets we simulated 1,000 observations. We run
2 models -- a two-distribution mixture model and a single-distribution
model -- each for both data sets. Models were run with 3 chains, with
each 6,000 iterations of which 3,000 were warmup. Estimates with 95\%
probability intervals are shown in Table \ref{tab:simparam}. The
parameters are shown by type of data generating process along with the
true parameter values. Parameter value estimates are shown by type of
Bayesian model. The results show that either of the two Bayesian models
succesfully uncovered the model parameters of the data with its
corresponding data generating process, as shown in Table
\ref{tab:simparam}, but less so when the model was applied to data
generated with the incorrect underlying process. Particularly the mixing
proportion \(\theta\) and the slowdown parameter \(\delta\) were not
uncovered at all by they single-distribution model.

@@

\begin{tabular}{lrr}
\toprule
 & \multicolumn{2}{c}{Estimate with 95\% PI} \\
\cmidrule(r){2-3}
Parameter with true value & \multicolumn{1}{c}{Two-distributions model} & \multicolumn{1}{c}{Single-distributions model}\\
\midrule
Two-distributions data &  & \\
\ \ \ $\beta = 5$ & 5.02 [4.99, 5.04] & 4.93 [4.76, 5.01]\\
\ \ \ $\delta = 1$ & 1.04 [0.92, 1.14] & 0.13 [0.01, 0.3]\\
\ \ \ $\theta = .35$ & .33 [.28, .39] & .54 [.11, .92]\\
\ \ \ $\sigma^2_1 = 0.25$ & 0.25 [0.23, 0.27] & 0.22 [0.16, 0.25]\\
\ \ \ $\sigma^2_2 = 0.5$ & 0.49 [0.42, 0.57] & 0.25 [0.22, 0.3]\\
Single-distribution data &  & \\
\ \ \ $\beta = 5$ & 5.35 [5.32, 5.39] & 5 [4.99, 5.02]\\
\ \ \ $\sigma = 0.25$ & 0.6 [0.57, 0.63] & 0.25 [0.24, 0.26]\\
\bottomrule
\end{tabular}

@@

We used LOO-CV to compare the predictive performance of the two models
for each data generating process. The model comparisons can be found for
each data generating process in Table \ref{tab:loossim}. The results
rule out the possibility that the mixture model does always lead to
higher predictive performance. Indeed, the mixture model showed a
slightly lower predictive performance for the data that were generated
with a single-distribution process. However, for the data generated with
a two-distribution mixture process, the mixture model shows a
substantially higher predictive performance. In fact, the ratio of
\(\Delta\widehat{elpd}\) and its standard error -- a \emph{z}-score
approximation -- reveals a value of 11.6 which is strong evidence in in
favour of the mixture model compared to the single-distribution model
(Sivula et al., 2020). In comparison, for the single-distribution data,
the single-distribution model was negligibly better than the
two-distribution mixture model which was evidence by a value of 0.77 for
the ratio of \(\Delta\widehat{elpd}\) and its standard error. Thus, as
the difference \(\Delta\widehat{elpd}\) between the model is negligible,
the single distribution is preferred by the law of parsimony.

@@

\begin{tabular}{lrr}
\toprule
Model & \multicolumn{1}{c}{$\Delta\widehat{elpd}$} & \multicolumn{1}{c}{$\widehat{elpd}$}\\
\midrule
Data: Two-distribution mixture process &  & \\
\ \ \ Two-distribution mixture model & -- & -6,068 (41)\\
\ \ \ Single-distribution model & -191.3 (16.5) & -6,259 (38)\\
Data: Single-distribution process &  & \\
\ \ \ Single-distribution model & -- & -5,030 (24)\\
\ \ \ Two-distribution mixture model & -0.5 (0.7) & -5,030 (24)\\
\bottomrule
\addlinespace
\end{tabular}
